Weak interspecific interactions in a sagebrush steppe: Evidence from observational data, population models, and a removal experiment

### 2016-08-17 removals/manuscript/removal_ms.tex
Edits on ms

\new{Here we present a removal experiment, conducted in a sagebrush steppe} 
% We conducted a removal experiment in a sagebrush steppe 
to test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release. We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses. We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs. removal treatments after accounting for differences in local species composition? Second, how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data?

@@

\subsection*{Removal experiment}
In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we selected locations for an additional 16 quadrats for the removal treatments. We used the following criteria to select the quadrat locations. Each plot had to have xx? cumulative cover/density? of the three target perennial grass species as well as yy? cover/density of \textit{Artemisia}. We mapped the quadrats in June, 2011 and then implemented the removal treatments in September (?), 2011. We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment, and the remaining 8 to the grass removal treatment. We cut sagebrush stems at ground level and applied herbicide to the cut stems. In the grass removal plots, we herbicided all perennial grasses. We removed plants zz cm beyond the quadrat boundaries to minimize edge effects. The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each additional year.

@@

Our regression models for survival, growth and recruitment follow previous work (Adler et al. 2010, Chu and Adler 2015) with a few exceptions, such as a new intercept term for the removal treatment. We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure), along with the removal treatment effect. Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
\begin{equation}
\mbox{logit}(S_{ijgh,t}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
\left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
\label{eqn:survReg}
\end{equation}
where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial group,$\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival. $\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species. The vector $\boldsymbol{W}$ includes crowding from the four dominant species,  \textit{A. tripartita}, \textit{P. spicata}, \textit{H. comata}, and \textit{Poa secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points. 
\new{$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
calculated as \texttt{sum(x*y)} in R.}

@@

\new{We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect 
than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
using the methods of Teller et al. (2016). The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
Demographic models such as \eqref{eqn:survReg} then have $\gamma$, $\varphi$, $\chi$ , 
$\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
by making the spline coefficients and $\eta$ the arguments of an objective function that computes $\boldsymbol{W}$ using the input spline coefficients, 
calls the model-fitting functions \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, 
and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
local minimum of AIC as a function of $df$, as in Teller et al. (2016). This approach assumes that one measure of crowding affects 
survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
within-species competition kernels $F_{jj}$. Based on previous work (Adler et al. 2010), we set all $F_{jm}$ equal to $F_{jj}$, meaning that 
the within-species competition kernel for species $j$ is also used to determine the effect of all other species on species $j$. We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}. }
\comment{SPE: see Robin's comment about whether you mean $F_{mm}$ or $F_{jj}$; I'm not sure which one you did.}

@@

We simulated the IPM to estimate equilibrium abundances under three different scenarios. The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects). We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium. To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of Metcalf et al. 2016))). We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period. For the second scenario, we set the cover of \textit{A. tripartita} to zero and then repeated the procedure. This simulation estimates the competitive release of grasses following \textit{A. tripartita} removal according to the baseline model (no removal effects). For the third scenario, we again set  \textit{A. triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects. By comparing the output of this simulation with the previous scenario we can quantify the long-term consequences of differences between our baseline model and a model that includes additional removal treatment effects.

@@

Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} 
(Fig. \ref{fig:CoverTrends}), but \new{shrub removal} did appear to cause an increase in the cover of \textit{P. spicata}. We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig. \ref{fig:CoverChange}). \new{Removal treatment} had no significant effect on 
\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.33, 0.27) or 
\textit{Poa secunda} (95\% CI: -0.40, 0.28) population growth, but did increase 
\textit{H. comata} (95\% CI: 0.02, 0.87) and
 \textit{P. spicata} (95\% CI: 0.04, 0.50) population growth 
(Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
adequately predicted by models fit to observational data on non-removal plots. 
\comment{SPE: I changed the CI numbers in this paragraph to match what I think the Table means.}

@@

Finally, we evaluated pre-existing differences in community composition in removal and control plots. In selecting removal plots, our criteria required candidate plots to include a certain amount of our target removal species. Because we adopted the historical plots as our controls, the same criteria were not applied (a control plot might not contain any \textit{A. tripartita}). To quantify resulting differences in pre-treatment (2011) composition, we compared neighborhood crowding at the individual plant level (the $\boldsymbol{W}$ of the growth regression) across treatments. \textit{A. tripartita} crowding and crowding by all subdominant shrub and grass species appeared greater than in the controls, whereas crowding by \textit{H. comata} was lower in removal plots than controls, though it was absent from most \textit{P. spicata} neighborhood in both treatments (Fig. \ref{fig:W-by-treatment}).

@@

\begin{table}[h]
\caption{Cover change models}
\centering
%\begin{center}
\comment{SPE: are the four columns actually the four species?}
\begin{tabular}{l c c c c }
\hline
 & Model 1 & Model 2 & Model 3 & Model 4 \\
Species & ARTR & HECO & POSE & PSSP \\
\hline
(Intercept)           & $-0.03$          & $-0.04$          & $-0.02$          & $-0.09$          \\
                      & $[-0.33;\ 0.27]$ & $[-0.32;\ 0.24]$ & $[-0.31;\ 0.27]$ & $[-0.33;\ 0.15]$ \\
Treatment: No grasses    & $-0.09$          &                  &                  &                  \\
                      & $[-0.45;\ 0.27]$ &                  &                  &                  \\
Treatment: No shrub    &                  & $0.44^{*}$       & $-0.06$          & $0.27^{*}$       \\
                      &                  & $[0.02;\ 0.87]$  & $[-0.40;\ 0.28]$ & $[0.04;\ 0.50]$  \\
\hline
AIC                   & 23.20            & 59.58            & 162.06           & 117.50           \\
BIC                   & 33.15            & 66.24            & 173.51           & 129.22           \\
Log Likelihood        & -6.60            & -24.79           & -76.03           & -53.75           \\
Num. obs.             & 54               & 28               & 73               & 77               \\
Num. groups: quad     & 12               & 5                & 14               & 14               \\
Num. groups: year     & 8                & 8                & 8                & 8                \\
Var: quad (Intercept) & 0.08             & 0.00             & 0.00             & 0.00             \\
Var: year (Intercept) & 0.01             & 0.00             & 0.10             & 0.09             \\
Var: Residual         & 0.04             & 0.32             & 0.40             & 0.19             \\
\hline
\multicolumn{5}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
\end{tabular}
\label{table:coefficients}
%\end{center}
\end{table}

>>>>

To test whether our models, which are based on observational data, underestimate the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe. We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses. We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs. removal treatments after accounting for differences in local species composition? Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic performance. If the full effect of experimental removals is stronger  than predicted, it means that our models underestimate competitive release.  Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data? If models based on observational data effectively predict competitive release, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments.

@@

\subsection*{Removal experiment}
In spring 2011, in a large exclosure containing six of the historical permanent quadrats, we selected locations for an additional 16 quadrats for the removal treatments. We used the following criteria to select the quadrat locations. Each plot had to have 5\% canopy cover of \textit{Artemisia tripartita} and at least 3\% combined basal cover of the three focal perennial grasses. We avoided plots falling on hill slopes, areas with greater than 20\% bare rock, or with over 10\% cover of the woody shrubs \textit{Purshia tridentata} or \textit{Amelanchier utahensis}. We mapped the quadrats in June, 2011 and then implemented the removal treatments in September, 2011. We randomly assigned 8 of the sixteen new quadrats to the sagebrush removal treatment, and the remaining 8 to the grass removal treatment. We cut sagebrush stems at ground level and applied a 5\% solution of Roundup Pro (Glyphosate, Monsanto) to the cut stems. In the grass removal plots, we painted the same herbicide solution on all perennial grasses. We removed plants 50 cm beyond the quadrat boundaries to minimize edge effects. The sagebrush removal was virtually 100\% effective; grass removals required additional herbicide applications in April and May of each year, with fewer grasses remaining each year.

@@

Our regression models for survival, growth and recruitment follow previous work (Adler et al. 2010, Chu and Adler 2015). We model the survival probability of an individual genet as a function of genet size, the neighborhood-scale crowding experienced by the genet from both conspecific and heterospecific genets, temporal variation among years, and permanent spatial variation among groups of quadrats (`group'; here means a set of nearby quadrats located within one pasture or grazing exclosure). Where the current analysis departs from previous work is the inclusion of a categorical removal treatment effect. If the conspecific and heterospecific crowding effects explain demographic rates in removal treatment plots as well as they do in control plots, then there will be no need for the removal treatment effect, and the estimated coefficient will be small and not significantly different from zero. However, if demographic rates in removal plots differ from control plots even after accounting for crowding effects, then the removal treatment effect will be significant. For example, we might find that, holding crowding constant, grasses grow more quickly in the removal plots than in the control plots. This result would indicate that our "baseline model," a model including all estimated parameters except for the removal treatment effect, underestimates competitive release. We refer to the model that includes the estimated removal treatment effect as the "removal model."

Formally, we modeled the survival probability, $S$, of genet $i$ in species $j$, group $g$, and removal treatment $h$, from time $t$ to $t+1$  as
\begin{equation}
\mbox{logit}(S_{ijgh,t}) = \gamma_{j,t}^S + \varphi_{jg}^S+  \chi_{jh}^S  + \beta_{j,t}^S u_{ij,t} +  
\left \langle \boldsymbol{\omega}_{j,t}^S, \boldsymbol{W}_{ij,t} \right \rangle 
\label{eqn:survReg}
\end{equation}
where $\gamma$ is a time-dependent intercept, $\varphi$ is the coefficient for the effect of spatial group, $\chi$ is the removal treatment effect, $\beta$ is the coefficient that represents the effect of log genet size, $u$, on survival. $\boldsymbol{\omega}$ is a vector of interaction coefficients which determine the impact of crowding, $\boldsymbol{W}$, by each species on the focal species. The vector $\boldsymbol{W}$ includes crowding from the four dominant species,  \textit{A. tripartita}, \textit{P. spicata}, \textit{H. comata}, and \textit{Poa secunda}, as well as two covariates representing total crowding from (1) all other perennial grasses and shrubs, which were mapped as polygons, and (2) all forb species, which were mapped as points. 
\new{$\left \langle \boldsymbol{x, y} \right \rangle$ denotes the inner product of vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, 
calculated as \texttt{sum(x*y)} in R.}

@@

We assume that competition kernels $F_{jm}(d)$ are non-negative and decreasing, so that distant plants have less effect 
than close plants. Otherwise, we let the data dictate the shape of the kernel by fitting a spline model 
using the methods of Teller et al. (2016). The shape of $F_{jm}$ is determined by a set of spline basis coefficients $\vec{b}_{jm}$
and a smoothing parameter $\eta$ that controls the complexity of the fitted kernel. 
Demographic models such as \eqref{eqn:survReg} then have $\gamma$, $\varphi$, $\chi$ , 
$\beta$, $\boldsymbol{\omega}$, $\vec{b}$ and $\eta$ as parameters to be fitted. We implemented this in the statistical computing environment, \texttt{R}, 
by making the spline coefficients and $\eta$ the arguments of an objective function that computes $\boldsymbol{W}$ using the input spline coefficients, 
calls the model-fitting functions \texttt{lmer} and/or \texttt{glmer} to fit the other parameters in the survival and growth regressions, 
and returns an approximate AIC value and model degrees of freedom ($df$) for survival and growth combined. We used the $\vec{b}$ values at the smoothest 
local minimum of AIC as a function of $df$, as in Teller et al. (2016). This approach assumes that one measure of crowding affects 
survival and growth. In addition, for fitting the kernels we assumed that survival and growth depended only on intraspecific crowding, and thus only fitted the
within-species competition kernels $F_{jj}$. Based on previous work (Adler et al. 2010), we set all $F_{mj}$ equal to $F_{jj}$, meaning that 
the within-species competition kernel for species $j$ is also used to determine the effect of all other species on species $j$. We used data from all historical plots and contemporary control-treatment plots to estimate the competition kernels, which are shown in Fig. \ref{fig:CompKernels}.

@@

We simulated the IPM to estimate equilibrium abundances under three different scenarios. The first scenario featured all four species and our ``baseline'' model for control plots (no removal effects). We initialized all species with arbitrarily low cover and ran the model for 500 time steps (years) to allow the community to reach an equilibrium. To incorporate temporal environmental variation, at each time step we randomly selected one of the 30 sets of random year effects for the survival, growth and recruitment models (``kernel selection'' in the terminology of Metcalf et al. 2016))). We calculated the equilibrium cover for each species by simulating an additional 2000 time steps and averaging each species' cover over this period. For the second scenario, we set the cover of \textit{A. tripartita} to zero and then repeated the procedure. This simulation estimates the competitive release of grasses following \textit{A. tripartita} removal according to the baseline model (no removal effects). For the third scenario, we again set  \textit{A. triparitia} to zero, but this time we projected the model using the vital rate regressions that include removal effects. If our baseline model successfully predicts competitive release observed in the removal experiment, then the results of the second and third scenarios should be similar. However, to the extent that the baseline model underestimates (or overestimates) the actual competitive release, the results of the second and third scenarios will differ.

@@

Trends in cover by species and removal treatment were variable. For example, grass removal had little effect on the cover of \textit{A. tripartita} 
(Fig. \ref{fig:CoverTrends}), but shrub removal did appear to cause an increase in the cover of \textit{P. spicata}. We tested the effects of removal on cover by analyzing year-to-year changes in (log) cover, a measure of population growth rate (Fig. \ref{fig:CoverChange}). The removal treatments had no significant effect on 
\textit{A. tripartita} (95\% confidence interval on removal treatment effect: -0.45, 0.27) or 
\textit{Poa secunda} (95\% CI: -0.40, 0.28) population growth, but did increase 
\textit{H. comata} (95\% CI: 0.02, 0.87) and
 \textit{P. spicata} (95\% CI: 0.04, 0.50) population growth 
(Table \ref{table:coefficients}). However, these treatment comparisons do not address our questions about whether responses to removal were 
adequately predicted by models fit to observational data on non-removal plots.

@@

Finally, we evaluated pre-existing differences in community composition in removal and control plots. In selecting removal plots, our criteria required candidate plots to include a certain amount of our target removal species. Because we adopted the historical plots as our controls, the same criteria were not applied (a control plot might not contain any \textit{A. tripartita}). To quantify resulting differences in pre-treatment (2011) composition, we compared neighborhood crowding at the individual plant level (the $\boldsymbol{W}$ of the growth regression) across treatments. \textit{A. tripartita} crowding and crowding by all subdominant shrub and grass species appeared greater than in the controls, whereas crowding by \textit{H. comata} was lower in removal plots than controls, though it was absent from most \textit{P. spicata} neighborhood in both treatments (Fig. \ref{fig:W-by-treatment}). These pre-existing differences in composition raise the possibility that plots selected for the removal treatments might differ from the control plots in some unobserved covariate that could influence plant performance.

@@

\begin{table}[h]
\caption{Cover change models}
\centering
%\begin{center}
\begin{tabular}{l c c c c }
\hline
Species & ARTR & HECO & POSE & PSSP \\
\hline
(Intercept)           & $-0.03$          & $-0.04$          & $-0.02$          & $-0.09$          \\
                      & $[-0.33;\ 0.27]$ & $[-0.32;\ 0.24]$ & $[-0.31;\ 0.27]$ & $[-0.33;\ 0.15]$ \\
Treatment: No grasses    & $-0.09$          &                  &                  &                  \\
                      & $[-0.45;\ 0.27]$ &                  &                  &                  \\
Treatment: No shrub    &                  & $0.44^{*}$       & $-0.06$          & $0.27^{*}$       \\
                      &                  & $[0.02;\ 0.87]$  & $[-0.40;\ 0.28]$ & $[0.04;\ 0.50]$  \\
\hline
AIC                   & 23.20            & 59.58            & 162.06           & 117.50           \\
BIC                   & 33.15            & 66.24            & 173.51           & 129.22           \\
Log Likelihood        & -6.60            & -24.79           & -76.03           & -53.75           \\
Num. obs.             & 54               & 28               & 73               & 77               \\
Num. groups: quad     & 12               & 5                & 14               & 14               \\
Num. groups: year     & 8                & 8                & 8                & 8                \\
Var: quad (Intercept) & 0.08             & 0.00             & 0.00             & 0.00             \\
Var: year (Intercept) & 0.01             & 0.00             & 0.10             & 0.09             \\
Var: Residual         & 0.04             & 0.32             & 0.40             & 0.19             \\
\hline
\multicolumn{5}{l}{\scriptsize{$^*$ 0 outside the confidence interval}}
\end{tabular}
\label{table:coefficients}
%\end{center}
\end{table}
